---
title: "Lab Report Knapsack"
author: "Kevin Duvert, Michael Nilsson"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Lab Report Knapsack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

\vspace{5mm}

We begin by running the code for generating the data that we will work with. 
\vspace{5mm}
```{r}
library(knapsack)
library(bench)
library(tidyverse)

set.seed(42)
n <- 2000
knapsack_objects <- data.frame(w = sample(1:4000, size = n, replace = TRUE),
                               v = runif(n = n, 0, 10000))
```
\vspace{5mm}

\normalsize
**Brute force search**

\vspace{5mm}
Below is our code for **brute_force_knapsack()**. We have implemented the function using **intToBits()**. It is important to note that because of the limitations of **intToBits()** we cannot choose arbitrarily large subsets.
\vspace{5mm}

```{r}
brute_force_knapsack<-function(x,W){
  if(!(is.data.frame(x)&ncol(x)==2&all(x>0))) stop("Incorrect input.")
  S<-c()
  L<-list()
  for (i in 1:(2^nrow(x))){
    if (sum(x[as.logical(intToBits(i-1)),][,1])>W) S[i]<-0
    else S[i]<-sum(x[as.logical(intToBits(i-1)),][,2])
    L[[i]]<-as.numeric(rownames(x[as.logical(intToBits(i-1)),]))
    L[[i]]<-as.numeric(rownames(x[as.logical(intToBits(i-1)),]))
  }
  list(value=max(S), elements=L[[which.max(S)]])
}
```
\vspace{1cm}
```{r}
brute_force_knapsack(x = knapsack_objects[1: 8,], W = 3500)
brute_force_knapsack(x = knapsack_objects[1:12,], W = 3500)
brute_force_knapsack(x = knapsack_objects[1:8,], W = 2000)
brute_force_knapsack(x = knapsack_objects[1:12,], W = 2000)
```
\vspace{1cm}
When trying the function for $n=16,$ we obtain the following results with **system.time()**:
\vspace{1cm}
```{r eval=FALSE}
bruteforce_st1<-system.time(brute_force_knapsack(x = knapsack_objects[1:16,], W = 3500))
bruteforce_st2<-system.time(brute_force_knapsack(x = knapsack_objects[1:16,], W = 2000))
```
```{r}
data("bruteforce_st1")
bruteforce_st1
data("bruteforce_st2")
bruteforce_st2
```

\vspace{2cm}

**Greedy heuristic**

\vspace{5mm}
Following the example provided in the instructions, we obtain the following code for **greedy_knapsack().**
\vspace{5mm}
```{r}
greedy_knapsack<-function(x,W){
  if(!(is.data.frame(x)&ncol(x)==2&all(x>0))) stop("Incorrect input.")
  x[,3]<-x[,2]/x[,1]
  x<-x[order(x[,3], decreasing=TRUE),]
  S<-0
  L<-c()
  s_w<-0
  for(i in 1:nrow(x)){
    if (s_w+x[i,1]<=W) {
      s_w<-s_w+x[i,1]
      S<-S+x[i,2]
      L<-c(L, as.numeric(rownames(x[i,])))
    }
  }
  list(value=S, elements=L)
}
```
\vspace{1cm}
```{r}
greedy_knapsack(x = knapsack_objects[1:800,], W = 3500)
greedy_knapsack(x = knapsack_objects[1:1200,], W = 2000)
```
\vspace{1cm}
When trying the function for $n=1000000,$ we obtain the following results with **system.time()**:
\vspace{1cm}
```{r eval=FALSE}
set.seed(42)
n2 <- 1000000
knapsack_objects2 <- data.frame(w = sample(1:4000, size = n2, replace = TRUE),
                               v <- runif(n = n2, 0, 10000))

greedy_st1<-system.time(greedy_knapsack(x = knapsack_objects2, W = 3500))
greedy_st2<-system.time(greedy_knapsack(x = knapsack_objects2, W = 2000))
```
```{r}
data("greedy_st1")
greedy_st1
data("greedy_st2")
greedy_st2
```

\vspace{2cm}
**Profiling and Optimization**

\vspace{5mm}
Using the package **profvis** we identified two bottlenecks for **brute_force_knapsack** and **greedy_knapsack** respectively; both arising inside for loops. We have tried our best to find alternative code that is more efficient. For **brute_force_knapsack** we replaced the for loop by using **sapply()** and **colSums()** (as suggested in Lecture 7). The alternative for **brute_force_knapsack** is given below:
\vspace{5mm}
```{r}
newbruteforce<-function(x,W){
 if(!(is.data.frame(x)&ncol(x)==2&all(x>0))) stop("Incorrect input.")
  y<-sapply(0:(2^(nrow(x))-1), function(k) {
     colSums(x[as.logical(intToBits(k)),])})
  y<-cbind(t(y),1:nrow(t(y)))
  y<-subset(y, y[,1]<=W)
  list(value=max(y[,2]),
       elements= as.numeric(rownames(x[as.logical(intToBits(y[which.max(y[,2]),3]-1)),])))
}
```
\vspace{5mm}
By using the **mark()** function from the **bench** package we can analyze performance differences between the two functions. We used the following code with 1000 iterations in order to make a comparison. The results are provided below (for the benchmark, **brute_force_knapsack** corresponds with the first row, and **newbruteforce** with the second). It seems like **newbruteforce** is generally faster than **brute_force_knapsack**.

\vspace{5mm}
```{r eval=FALSE}
bruteforcebench<-bench::mark(brute_force_knapsack(x = knapsack_objects[1:8,], W = 3500),
            newbruteforce(x = knapsack_objects[1:8,], W = 3500), iterations=1000)
```
```{r}
data("bruteforcebench")
bruteforcebench[,-1]
```

\vspace{5mm}
For **greedy_knapsack()** we were able to adapt the code so that the function implements the for loop while treating the data frame as a matrix. The code is the following one:
\vspace{5mm}
```{r}
newgreedy<-function(x,W){
  x<-as.matrix(x)
  x<-cbind(x,x[,2]/x[,1],1:nrow(x))
  x<-x[order(x[,3], decreasing=TRUE),]
  S<-0
  L<-c()
  s_w<-0
  for(i in 1:nrow(x)){
    if (s_w+x[i,1]<=W) {
      s_w<-s_w+x[i,1]
      S<-S+x[i,2]
      L<-c(L,x[i,4])
    }
  }
  list(value=S, elements=L)
}
```
\vspace{5mm}
We compared **greedy_knapsack** and **newgreedy** using **mark()** (1000 iterations). Here the result seems staggering. **newgreedy** seemed to vastly outperform **greedy_knapsack** in execution time. The code that was used is the one below (for the benchmark, **greedy_knapsack** corresponds with the first row, and **newgreedy** with the second):
\vspace{5mm}
```{r eval=FALSE}
greedybench<-bench::mark(greedy_knapsack(x = knapsack_objects[1:800,], W = 3500),
                         newgreedy(x = knapsack_objects[1:800,], W = 3500), iterations=1000)
```
```{r}
data(greedybench)
greedybench[,-1]
```

\vspace{5mm}
In our package, both **brute_force_knapsack** and **greedy_knapsack** are implemented using the new code. 
